---
title: "geneOptimize: An Object-Oriented Framework for Genetic Algorithms and Variable Selection in R"
author: "Hui Zheng"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{geneOptimize: An Object-Oriented Framework for Genetic Algorithms and Variable Selection in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  eval      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Abstract

**geneOptimize** is an R package that provides a modular, object-oriented
framework for genetic algorithms (GAs) and memetic algorithms.  Built on the
R6 class system, it supports binary and real-valued encodings, plug-in
operator modules (selection, crossover, mutation), parallel fitness
evaluation, and hybridization with local search (memetic algorithms).  Version
1.1.0 introduces input validation with informative error messages, a dedicated
`maximize` parameter for minimization problems, explicit `lower`/`upper`
bound arguments for real-valued optimization, selective memetic local search
controls, vectorized diversity tracking, a corrected `MutationGaussian`
interface, a fully functional `Population` class, and a comprehensive
`testthat` test suite.

---

# Introduction

Genetic algorithms are population-based meta-heuristics inspired by
Darwinian natural selection [@Holland1975].  At each generation a GA evaluates
a population of candidate solutions (chromosomes), selects the fitter
individuals, and applies genetic operators (crossover, mutation) to produce
offspring.  Over many generations the population converges toward high-quality
solutions.

GAs are widely applied to problems that are non-convex, non-differentiable, or
combinatorially complex — including feature selection for statistical models
[@Guyon2003], hyperparameter optimization [@Bergstra2012], and real-valued
function optimization [@Goldberg1989].

**geneOptimize** provides a clean R6-based framework that separates concerns
into:

- **Chromosome classes** — encoding and mutation logic.
- **Population class** — initialization and evaluation.
- **Operator classes** — modular, swappable selection/crossover/mutation.
- **`run_ga()` functional interface** — high-level entry point.
- **Plotting suite** — rich convergence diagnostics.

---

# Installation

```{r install, eval=FALSE}
# From CRAN
install.packages("geneOptimize")

# Development version from GitHub
# install.packages("remotes")
remotes::install_github("huizheng1-lab/geneOptimize")
```

---

# Architecture

## Class Hierarchy

```
Chromosome (abstract)
  ├── BinaryChromosome   — bit-vector encoding, bit-flip mutation
  └── RealChromosome     — real-vector encoding, Gaussian noise mutation

Population             — list of Chromosomes + evaluate() / get_best()
GeneticAlgorithm       — orchestration wrapper (delegates to run_ga())

SelectionOperator (abstract)
  ├── SelectionTournament  — tournament selection (configurable k)
  └── SelectionRoulette    — fitness-proportionate selection

CrossoverOperator (abstract)
  ├── CrossoverSinglePoint — single-point crossover
  ├── CrossoverUniform     — uniform crossover (random mask)
  └── CrossoverArithmetic  — arithmetic blend (real-valued)

MutationOperator (abstract)
  ├── MutationSimple    — bit-flip (binary) or scaled Gaussian (real)
  └── MutationGaussian  — Gaussian noise with configurable sigma
```

The strategy pattern is used throughout: any operator can be replaced by a
user-defined R6 class that inherits from the appropriate base class and
implements the required method (`select()`, `mate()`, or `mutate()`).

## Main Entry Point

The `run_ga()` function is the primary user-facing interface.  It accepts an
R6 operator for each GA phase and orchestrates the full evolutionary loop.

---

# Functional Interface: `run_ga()`

## Signature

```{r run_ga_sig}
run_ga(
  fitness_fn,                              # Required
  n_genes,                                 # Required
  pop_size           = 50,
  generations        = 100,
  selection_fn       = SelectionTournament$new(),
  crossover_fn       = CrossoverSinglePoint$new(),
  mutation_fn        = MutationSimple$new(),
  crossover_rate     = 0.8,
  mutation_rate      = 0.01,
  type               = "binary",           # "binary" or "real"
  lower              = NULL,               # required when type = "real"
  upper              = NULL,               # required when type = "real"
  maximize           = TRUE,               # FALSE for minimization
  elitism_count      = 1,
  verbose            = TRUE,
  parallel           = FALSE,
  cores              = 1,
  local_search_fn    = NULL,
  local_search_every = 1,
  local_search_top_k = NULL,
  track_diversity    = FALSE,
  ...
)
```

## Algorithm

1. **Initialize** population (binary: random 0/1; real: uniform over
   `[lower, upper]`).
2. **For each generation:**
   a. Evaluate fitness (optionally in parallel).
   b. Record `best_history`, `mean_history`, `worst_history`; optionally
      compute `diversity_history` using `stats::dist()`.
   c. Apply **elitism**: copy the top `elitism_count` individuals unchanged.
   d. **Selection** → build a mating pool.
   e. **Crossover** → generate offspring with probability `crossover_rate`.
   f. **Mutation** → perturb offspring gene-by-gene.
   g. **Local search** (optional): apply `local_search_fn` every
      `local_search_every` generations, to all or the top-`k` individuals.
   h. Restore elite individuals (they override the first rows).
3. **Return** a `gene_opt_res` object.

## Maximization and Minimization

Prior to v1.1.0, `run_ga()` always maximized the fitness function; users had
to negate their objective to minimize.  Version 1.1.0 adds:

```{r maximize_param}
# Minimize the Sphere function
result <- run_ga(
  fitness_fn  = function(x) sum(x^2),
  n_genes     = 5,
  type        = "real",
  lower       = rep(-5, 5),
  upper       = rep( 5, 5),
  maximize    = FALSE,          # <-- new in v1.1.0
  pop_size    = 50,
  generations = 100,
  verbose     = FALSE
)
```

## Explicit Bounds for Real-Valued Optimization

`lower` and `upper` are now **named parameters** in the function signature
(previously they had to be passed via `...`), which makes the interface
clearer and enables proper validation:

```{r bounds_example}
result <- run_ga(
  fitness_fn  = function(x) -sum((x - 2)^2),
  n_genes     = 3,
  type        = "real",
  lower       = c(-5, -5, -5),    # explicit lower bounds
  upper       = c( 5,  5,  5),    # explicit upper bounds
  pop_size    = 40,
  generations = 80,
  verbose     = FALSE
)
```

Scalar bounds are recycled to length `n_genes`:

```{r scalar_bounds}
result <- run_ga(..., lower = -10, upper = 10)
```

## Input Validation

All parameters are validated at startup with informative error messages:

```{r validation_demo, error=TRUE}
run_ga(42, n_genes = 5)
#> Error: 'fitness_fn' must be a function.

run_ga(sum, n_genes = 5, type = "real")
#> Error: 'lower' and 'upper' must be provided when type = "real".

run_ga(sum, n_genes = 5, crossover_rate = 1.5)
#> Error: 'crossover_rate' must be in [0, 1].
```

---

# Operators

## Selection

### Tournament Selection

Randomly draws `k` individuals and returns the one with the highest
(or lowest, for minimization) fitness.  Default `k = 3`.

```{r tournament}
sel <- SelectionTournament$new(k = 5)
```

### Roulette-Wheel Selection

Fitness-proportionate selection.  In v1.1.0 the shift offset was changed from
the ad-hoc `0.01` to `1e-6` and is now documented in the source code.
The edge case of all-equal fitness values (previously undefined behaviour) now
correctly produces uniform random selection:

```{r roulette}
sel <- SelectionRoulette$new()
```

## Crossover

| Class | Description |
|-------|-------------|
| `CrossoverSinglePoint` | Swap gene segments at one random point |
| `CrossoverUniform` | Each gene independently inherited from either parent |
| `CrossoverArithmetic` | Weighted blend: `c1 = α·p1 + (1−α)·p2` |

## Mutation

### MutationSimple

Handles both encodings:

- **Binary** — bit-flip with probability `mutation_rate`.
- **Real** — additive Gaussian noise scaled by `(upper − lower) / 10`.

### MutationGaussian (revised in v1.1.0)

The `MutationGaussian` class previously had an inconsistent method signature
`mutate(genes, rate, sigma, lower, upper)` that did not match the interface
expected by `run_ga()`.  In v1.1.0 its signature is aligned with
`MutationSimple`:

```{r gaussian_mutation}
# sigma is now a class field set at construction time
mut <- MutationGaussian$new(sigma = 0.05)

result <- run_ga(
  fitness_fn  = function(x) -sum(x^2),
  n_genes     = 4,
  type        = "real",
  lower       = rep(-1, 4),
  upper       = rep( 1, 4),
  mutation_fn = mut,
  pop_size    = 30,
  generations = 50,
  verbose     = FALSE
)
```

---

# Binary Encoding: Variable Selection

Variable selection for statistical models is a natural GA application:
a binary chromosome of length $p$ encodes which of the $p$ candidate features
are included.

```{r var_selection}
library(geneOptimize)

set.seed(42)
n <- 200; p <- 15
X <- matrix(rnorm(n * p), n, p)
colnames(X) <- paste0("V", seq_len(p))

# Ground truth: only V1, V3, V5 affect outcome
logits <- 1.5 * X[, 1] - 1.2 * X[, 3] + 0.8 * X[, 5]
y <- as.factor(ifelse(runif(n) < plogis(logits), "Class1", "Class0"))
df <- data.frame(y = y, X)

fitness_sparse_auc <- function(genes, penalty = 0.01) {
  if (sum(genes) == 0) return(0)
  vars  <- colnames(X)[genes == 1]
  model <- glm(as.formula(paste("y ~", paste(vars, collapse = " + "))),
               data = df, family = binomial)
  preds   <- predict(model, type = "response")
  roc_obj <- pROC::roc(y, preds, quiet = TRUE)
  as.numeric(pROC::auc(roc_obj)) - penalty * sum(genes)
}

result <- run_ga(
  fitness_fn  = fitness_sparse_auc,
  n_genes     = p,
  type        = "binary",
  pop_size    = 50,
  generations = 40,
  verbose     = FALSE
)

print(colnames(X)[result$best_chromosome == 1])
```

---

# Real-Valued Optimization

## Ackley Function

The Ackley function is a standard multi-modal benchmark with a global
minimum of 0 at the origin [@Ackley1987].

```{r ackley}
ackley <- function(x) {
  d    <- length(x)
  a <- 20; b <- 0.2; c <- 2 * pi
  -(- a * exp(-b * sqrt(sum(x^2) / d))
    - exp(sum(cos(c * x)) / d)
    + a + exp(1))
}

result <- run_ga(
  fitness_fn  = ackley,
  n_genes     = 2,
  type        = "real",
  lower       = rep(-5, 2),
  upper       = rep( 5, 2),
  maximize    = FALSE,
  mutation_fn = MutationGaussian$new(sigma = 0.1),
  crossover_fn = CrossoverArithmetic$new(alpha = 0.5),
  pop_size    = 50,
  generations = 100,
  verbose     = FALSE
)
print(result)
```

---

# Memetic Algorithms

Memetic algorithms [@Moscato1989] hybridize evolutionary search with local
refinement.  `run_ga()` supports this through three parameters new or
improved in v1.1.0:

| Parameter | Type | Description |
|-----------|------|-------------|
| `local_search_fn` | function | Takes a chromosome, returns improved chromosome |
| `local_search_every` | integer | Apply every _n_ generations (default: 1) |
| `local_search_top_k` | integer / NULL | Apply only to top-_k_ individuals |

**New in v1.1.0:** `local_search_every` and `local_search_top_k` allow
fine-grained control over the computational cost of local search.

```{r memetic}
fitness_fn <- function(x) -sum((x - 0.5)^2)

# Hill-climbing local search
local_search <- function(chromosome) {
  candidate <- chromosome + rnorm(length(chromosome), 0, 0.01)
  if (fitness_fn(candidate) > fitness_fn(chromosome)) candidate else chromosome
}

result <- run_ga(
  fitness_fn         = fitness_fn,
  n_genes            = 5,
  type               = "real",
  lower              = rep(0, 5),
  upper              = rep(1, 5),
  pop_size           = 30,
  generations        = 50,
  local_search_fn    = local_search,
  local_search_every = 5,       # apply every 5 generations
  local_search_top_k = 5,       # apply only to the 5 best individuals
  verbose            = FALSE
)
```

---

# Parallel Processing

`run_ga()` parallelizes fitness evaluation across multiple CPU cores.  On
Unix-like systems `parallel::mclapply()` is used; on Windows a socket cluster
via `parallel::makeCluster()` is created and cleaned up automatically.

**Requirements for `parallel = TRUE`:**

- `fitness_fn` must be a *pure function*: no global-state reads/writes, no
  side effects.
- All objects needed by `fitness_fn` must be captured in its closure.

```{r parallel}
result <- run_ga(
  fitness_fn  = function(x) sum(x),
  n_genes     = 20,
  pop_size    = 100,
  generations = 200,
  parallel    = TRUE,
  cores       = 4,
  verbose     = FALSE
)
```

---

# Visualization

## Convergence Curve

```{r plot_conv}
plot_convergence(result)
```

## Fitness Statistics (Best / Mean / Worst)

```{r plot_stats}
plot_fitness_stats(result)
```

## Population Diversity

Diversity is computed as mean pairwise distance (Hamming for binary, Euclidean
for real) using the vectorized `stats::dist()` function — replacing the
slower `Vectorize(outer())` loop from v1.0.0.

**Note:** diversity tracking is disabled by default (`track_diversity = FALSE`)
because the O(n²) computation is non-trivial for large populations.  The
error message raised by `plot_diversity()` when diversity was not tracked now
includes the exact corrective action:

```{r plot_div}
result_div <- run_ga(
  fitness_fn      = function(x) sum(x),
  n_genes         = 10,
  pop_size        = 30,
  generations     = 50,
  track_diversity = TRUE,
  verbose         = FALSE
)
plot_diversity(result_div)
```

## Diagnostic Panel

```{r plot_diag}
# conv_threshold is now user-configurable (default: 0.001 = 0.1%)
plot_diagnostics(result, conv_threshold = 0.005)
```

## Comparing Runs

```{r plot_cmp}
plot_compare(result1, result2, result3,
             labels = c("Tournament", "Roulette", "Uniform Crossover"))
```

---

# Population and GeneticAlgorithm Classes

Version 1.1.0 completes the OOP infrastructure that was partially stubbed in
v1.0.0.

## Population

```{r population_ex}
pop <- Population$new(size = 20, n_genes = 10, type = "binary")
pop$evaluate(function(x) sum(x))
best <- pop$get_best()
cat("Best fitness:", best$fitness, "\n")

# Real-valued population
pop_real <- Population$new(size = 20, n_genes = 5, type = "real",
                           lower = -1, upper = 1)
```

## GeneticAlgorithm

```{r ga_class_ex}
pop    <- Population$new(size = 30, n_genes = 10, type = "binary")
ga_obj <- GeneticAlgorithm$new(
  pop           = pop,
  fitness_fn    = function(x) sum(x),
  generations   = 50,
  crossover_rate = 0.8,
  mutation_rate  = 0.01
)
result <- ga_obj$run(verbose = FALSE)
plot_convergence(result)
```

---

# Testing

Version 1.1.0 introduces a full `testthat` [@Wickham2011] test suite covering:

| Test file | Coverage |
|-----------|----------|
| `test-chromosome.R` | `BinaryChromosome`, `RealChromosome` |
| `test-operators.R` | All selection, crossover, and mutation operators |
| `test-run_ga.R` | Input validation, binary/real types, maximize, elitism, diversity, local search |
| `test-methods.R` | `new_gene_opt_res`, `print`, `summary`, `plot` |
| `test-plotting.R` | All plotting functions, error conditions |
| `test-population.R` | `Population` initialization, evaluate, get_best |

Run the test suite with:

```{r test_suite, eval=FALSE}
testthat::test_package("geneOptimize")
# or
devtools::test()
```

---

# Custom Operators

Users can extend any operator class to implement custom behaviour:

```{r custom_op}
library(R6)

# Custom selection: epsilon-greedy (exploit best with probability 1-eps,
# explore randomly with probability eps)
EpsilonGreedySelection <- R6Class(
  "EpsilonGreedySelection",
  inherit = SelectionOperator,
  public = list(
    eps = 0.1,
    initialize = function(eps = 0.1) { self$eps <- eps },
    select = function(population, fitness_values, n = 1, ...) {
      if (runif(1) < self$eps) {
        return(population[sample(nrow(population), 1), ])
      }
      population[which.max(fitness_values), ]
    }
  )
)

result <- run_ga(
  fitness_fn   = function(x) sum(x),
  n_genes      = 15,
  selection_fn = EpsilonGreedySelection$new(eps = 0.2),
  pop_size     = 40,
  generations  = 60,
  verbose      = FALSE
)
```

---

# Summary of Changes in v1.1.0

| Category | Change |
|----------|--------|
| `run_ga()` | Added `maximize`, `lower`, `upper` as explicit named parameters |
| `run_ga()` | Comprehensive input validation with informative error messages |
| `run_ga()` | Added `local_search_every` and `local_search_top_k` for selective memetic local search |
| `run_ga()` | Replaced `Vectorize(outer())` diversity with `stats::dist()` |
| `run_ga()` | Documented parallel-mode requirements in Rd documentation |
| `MutationGaussian` | Fixed method signature to match `MutationSimple` interface; `sigma` is now a class field |
| `SelectionRoulette` | Replaced magic offset `0.01` with documented `1e-6`; handles all-equal fitness |
| `plot_diagnostics()` | Added user-configurable `conv_threshold` parameter |
| `plot_diversity()` | Error message now includes actionable corrective instruction |
| `Population` | Completed `initialize()`, `evaluate()`, and `get_best()` |
| `GeneticAlgorithm` | Completed `run()` method (delegates to `run_ga()`) |
| Tests | Added full `testthat` suite (60+ tests across 6 files) |

---

# Conclusion

**geneOptimize** provides a production-ready, extensible framework for genetic
and memetic algorithms in R.  Its R6-based architecture enables easy operator
swapping and extension, while the `run_ga()` functional interface keeps common
use cases concise.  Version 1.1.0 focuses on correctness, usability, and
testability — making the package suitable for both research prototyping and
production variable selection pipelines.

---

# References
